---
title: "NYC Rent and Subway Proximity: Exploratory Analysis"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
knitr::opts_knit$set(root.dir = normalizePath(".."))

library(tidyverse)

theme_set(theme_minimal())
```

## Purpose

This analysis examines the quality, distributions, and relationships in the NYC rental and subway data before regression modeling. We use only the exploration sample so that the confirmation sample remains separate until the final analysis.

## Load exploration data

```{r load-data}
analysis_data <- read.csv(
  file.path("data", "processed", "nyc_rent_transit.csv"),
  na.strings = c("", "NA")
)

eda_df <- analysis_data |>
  filter(analysis_split == "exploration")
```

## Dataset overview

Each row represents one advertised rental listing. Monthly rent is the outcome, and distance to the nearest subway station is the main variable of interest.

```{r dataset-overview}
dim(eda_df)
glimpse(eda_df)
```

## Data quality

### Impossible zero values

Check whether zero represents a possible value or missing information.

```{r zero-values}
zero_summary <- tibble(
  variable = c("Square footage", "Net effective rent", "Months free"),
  zero_count = c(
    sum(eda_df$sqft == 0, na.rm = TRUE),
    sum(eda_df$net_effective_rent == 0, na.rm = TRUE),
    sum(eda_df$months_free == 0, na.rm = TRUE)
  )
) |>
  mutate(zero_percent = round(100 * zero_count / nrow(eda_df), 1))

knitr::kable(
  zero_summary,
  col.names = c("Variable", "Zero count", "Zero percent"),
  caption = "Zero values in the exploration sample"
)
```

```{r sqft-zero-distribution, fig.width=7, fig.height=4}
ggplot(eda_df, aes(x = sqft)) +
  geom_histogram(
    binwidth = 100,
    boundary = 0,
    color = "white",
    fill = "#2C7FB8"
  ) +
  coord_cartesian(xlim = c(0, 3000)) +
  labs(
    title = "Reported square footage",
    subtitle = "Exploration sample",
    x = "Square feet",
    y = "Number of listings"
  )
```

Zero square footage and zero net-effective rent are not possible, so we treat them as missing. Zero free months is meaningful and remains unchanged.

```{r recode-zero-values}
eda_df <- eda_df |>
  mutate(
    sqft = na_if(sqft, 0),
    net_effective_rent = na_if(net_effective_rent, 0)
  )
```

### Missing values

```{r missing-values}
missing_summary <- eda_df |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_count"
  ) |>
  mutate(missing_percent = round(100 * missing_count / nrow(eda_df), 1)) |>
  filter(missing_count > 0) |>
  arrange(desc(missing_percent))

knitr::kable(
  missing_summary,
  col.names = c("Variable", "Missing count", "Missing percent"),
  caption = "Missing values in the exploration sample"
)
```

Net-effective rent, lease length, and square footage have the most missing values. Using them as required model variables would remove many listings.

### Possible duplicates

Repeated listings could give the same apartment extra weight and make the observations less independent. First, count repeated IDs, URLs, and known address-unit combinations.

```{r duplicates}
eda_df |>
  summarise(
    duplicate_listing_ids = sum(duplicated(listing_id)),
    duplicate_listing_urls = sum(duplicated(listing_url)),
    duplicate_address_units = sum(
      !is.na(unit) & duplicated(paste(borough, street, unit))
    )
  )
```

Listing IDs are unique, but some URLs and address-unit combinations appear more than once. Look at those listings more closely.

```{r duplicate-details}
repeated_urls <- eda_df |>
  add_count(listing_url, name = "times_listed") |>
  filter(times_listed > 1) |>
  arrange(listing_url) |>
  select(
    listing_id, listing_url, borough, street, unit,
    monthly_rent, available_date, times_listed
  )

repeated_units <- eda_df |>
  filter(!is.na(unit)) |>
  add_count(borough, street, unit, name = "times_listed") |>
  filter(times_listed > 1) |>
  arrange(borough, street, unit) |>
  select(
    listing_id, listing_url, borough, street, unit,
    monthly_rent, available_date, times_listed
  )

repeated_urls
repeated_units
```

Many of these records appear to be apartment relistings. We will keep one row per listing URL and one row per borough, street, and unit combination. Since there is no reliable listing date, we will keep the first record in the dataset. This rule is applied to both samples in `src/data/04_prepare_modeling_data.R`.

## Core variables

Summarize rent and subway distance overall and by borough.

```{r core-summary}
core_summary <- eda_df |>
  summarise(
    listings = n(),
    median_rent = median(monthly_rent, na.rm = TRUE),
    mean_rent = mean(monthly_rent, na.rm = TRUE),
    sd_rent = sd(monthly_rent, na.rm = TRUE),
    minimum_rent = min(monthly_rent, na.rm = TRUE),
    maximum_rent = max(monthly_rent, na.rm = TRUE),
    median_distance_m = median(distance_to_subway_m, na.rm = TRUE),
    mean_distance_m = mean(distance_to_subway_m, na.rm = TRUE),
    sd_distance_m = sd(distance_to_subway_m, na.rm = TRUE),
    minimum_distance_m = min(distance_to_subway_m, na.rm = TRUE),
    maximum_distance_m = max(distance_to_subway_m, na.rm = TRUE)
  )

knitr::kable(
  core_summary,
  digits = 1,
  caption = "Summary of rent and subway distance"
)
```

```{r borough-summary}
borough_summary <- eda_df |>
  group_by(borough) |>
  summarise(
    listings = n(),
    median_rent = median(monthly_rent, na.rm = TRUE),
    mean_rent = mean(monthly_rent, na.rm = TRUE),
    median_distance_m = median(distance_to_subway_m, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(median_rent))

knitr::kable(
  borough_summary,
  digits = 1,
  caption = "Rent and subway distance by borough"
)
```

The median listing has a monthly rent of $4,000 and is 309 meters from the nearest subway station. Manhattan has the highest median rent and the shortest median subway distance. Staten Island has very few exploration listings, so its summaries should be interpreted cautiously.

## Distributions

### Monthly rent

Start by viewing monthly rent in its original units.

```{r rent-distribution, fig.width=7, fig.height=5}
ggplot(eda_df, aes(x = monthly_rent)) +
  geom_histogram(bins = 40, color = "white", fill = "#2C7FB8") +
  geom_vline(
    xintercept = median(eda_df$monthly_rent, na.rm = TRUE),
    color = "darkgreen",
    linewidth = 1,
    linetype = "dashed"
  ) +
  geom_vline(
    xintercept = mean(eda_df$monthly_rent, na.rm = TRUE),
    color = "firebrick",
    linewidth = 1,
    linetype = "dashed"
  ) +
  scale_x_continuous(labels = scales::label_dollar()) +
  labs(
    title = "Distribution of advertised monthly rent",
    subtitle = "Median in green and mean in red",
    x = "Monthly rent",
    y = "Number of listings"
  )
```

Monthly rent is right-skewed, with a smaller number of high-priced listings forming a long upper tail. The mean exceeds the median because these listings pull the average upward.

```{r rent-log-distribution, fig.width=7, fig.height=5}
ggplot(eda_df, aes(x = monthly_rent)) +
  geom_histogram(bins = 40, color = "white", fill = "#2C7FB8") +
  scale_x_log10(labels = scales::label_dollar()) +
  labs(
    title = "Distribution of advertised monthly rent on a log scale",
    subtitle = "Exploration sample",
    x = "Monthly rent, shown on a log scale",
    y = "Number of listings"
  )
```

### Distance to the nearest subway station

```{r distance-distribution, fig.width=7, fig.height=5}
ggplot(eda_df, aes(x = distance_to_subway_m)) +
  geom_histogram(bins = 40, color = "white", fill = "#2C7FB8") +
  geom_vline(
    xintercept = median(eda_df$distance_to_subway_m, na.rm = TRUE),
    color = "darkgreen",
    linewidth = 1,
    linetype = "dashed"
  ) +
  geom_vline(
    xintercept = mean(eda_df$distance_to_subway_m, na.rm = TRUE),
    color = "firebrick",
    linewidth = 1,
    linetype = "dashed"
  ) +
  labs(
    title = "Distance to the nearest subway station",
    subtitle = "Median in green and mean in red",
    x = "Distance in meters",
    y = "Number of listings"
  )
```

Subway distance is also right-skewed. Most listings are relatively close to a station, while a small number are much farther away.

```{r distance-log-distribution, fig.width=7, fig.height=5}
ggplot(eda_df, aes(x = distance_to_subway_m)) +
  geom_histogram(bins = 40, color = "white", fill = "#2C7FB8") +
  scale_x_log10(labels = scales::label_number()) +
  labs(
    title = "Distribution of subway distance on a log scale",
    subtitle = "Exploration sample",
    x = "Distance in meters, shown on a log scale",
    y = "Number of listings"
  )
```

The log-scaled distributions are less right-skewed. We will compare models with and without these transformations.

## Rent and subway distance

First, view the relationship in its original units.

```{r rent-distance-original, fig.width=7, fig.height=5}
ggplot(
  eda_df,
  aes(x = distance_to_subway_m, y = monthly_rent)
) +
  geom_point(alpha = 0.08, size = 0.8, color = "#2C7FB8") +
  geom_smooth(method = "lm", se = TRUE, color = "firebrick") +
  scale_y_continuous(labels = scales::label_dollar()) +
  labs(
    title = "Advertised rent and distance to the nearest station",
    subtitle = "Exploration sample",
    x = "Distance in meters",
    y = "Monthly rent"
  )
```

The plot suggests a modest negative relationship, but rent varies substantially at similar subway distances.

```{r rent-distance-logged, fig.width=7, fig.height=5}
ggplot(
  eda_df,
  aes(x = distance_to_subway_m, y = monthly_rent)
) +
  geom_point(alpha = 0.08, size = 0.8, color = "#2C7FB8") +
  geom_smooth(method = "lm", se = TRUE, color = "firebrick") +
  scale_x_log10() +
  scale_y_log10(labels = scales::label_dollar()) +
  labs(
    title = "Advertised rent and subway distance on log scales",
    subtitle = "Exploration sample",
    x = "Distance in meters, shown on a log scale",
    y = "Monthly rent, shown on a log scale"
  )
```

## Correlations

Use pairwise correlations to compare the linear relationships among the main numeric variables. Each value uses the rows where both variables are available.

```{r correlation-matrix, fig.width=9, fig.height=8}
correlation_data <- eda_df |>
  select(
    monthly_rent,
    net_effective_rent,
    lease_months,
    distance_to_subway_m,
    bedrooms,
    bathrooms,
    half_baths,
    sqft,
    months_free
  ) |>
  rename(
    Rent = monthly_rent,
    `Net effective rent` = net_effective_rent,
    `Lease months` = lease_months,
    `Subway distance` = distance_to_subway_m,
    Bedrooms = bedrooms,
    Bathrooms = bathrooms,
    `Half baths` = half_baths,
    `Square feet` = sqft,
    `Free months` = months_free
  )

correlation_matrix <- cor(
  correlation_data,
  use = "pairwise.complete.obs"
)

correlation_plot_data <- as.data.frame(as.table(correlation_matrix)) |>
  rename(variable_x = Var1, variable_y = Var2, correlation = Freq)

ggplot(
  correlation_plot_data,
  aes(x = variable_x, y = variable_y, fill = correlation)
) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(correlation, 2)), size = 3) +
  scale_fill_gradient2(
    low = "#B2182B",
    mid = "white",
    high = "#2166AC",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    title = "Correlations among numeric variables",
    x = NULL,
    y = NULL,
    fill = "Correlation"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Rent is most strongly correlated with net-effective rent, bathrooms, square footage, and bedrooms. Its correlation with subway distance is weak and negative. Correlations involving net-effective rent, lease length, or square footage use much smaller samples because these variables are missing for many listings.

## Monthly rent by borough

```{r rent-by-borough, fig.width=8, fig.height=5}
ggplot(
  eda_df,
  aes(
    x = reorder(borough, monthly_rent, FUN = median),
    y = monthly_rent
  )
) +
  geom_boxplot(fill = "#2C7FB8", outlier.alpha = 0.08) +
  scale_y_continuous(labels = scales::label_dollar()) +
  coord_flip() +
  labs(
    title = "Monthly rent varies by borough",
    subtitle = "Exploration sample",
    x = NULL,
    y = "Monthly rent"
  )
```

Manhattan has the highest median monthly rent, followed by Brooklyn. Borough differences help explain why location should be included in the fuller descriptive model.

## Impact of EDA

Decisions from this analysis are implemented in `src/data/04_prepare_modeling_data.R`. The script currently recodes impossible zero values, removes repeated listings, and creates modeling features. It can be updated as more decisions are made.
