---
title: "NYC Rent and Subway Proximity: Modelling Exploration"
output: html_document
---

#Install libraries and load modeling data

```{r}
library(tidyverse)
library(sandwich)
library(lmtest)
library(stargazer)

modeling_data <- read.csv(
  file.path("data", "processed", "nyc_rent_transit_modeling.csv"),
  na.strings = c("", "NA")
)
```
# Set up dataframes

```{r}
# Full exploration sample
explore_df <- modeling_data |>
  filter(analysis_split == "exploration")

# Complete-case sample for models requiring all core covariates
model_df <- explore_df |>
  drop_na(
    monthly_rent,
    bedrooms,
    bathrooms,
    sqft,
    borough,
    distance_to_subway_m
  )

nrow(explore_df)
nrow(model_df)
```
# Run m1

This simple model m1 just captures the unconditional relationship between our primary X (distance to transit stop) and Y (monthly rent) variables. We'll run the model on both our exploratory dataframe (which keeps all the rows) and our filtered dataframe that drops rows with missing variable values. 

```{r}
m1 <- lm(
  monthly_rent ~ distance_to_subway_m,
  data = explore_df
)

m1_common <- lm(
  monthly_rent ~ distance_to_subway_m,
  data = model_df
)

summary(m1)
summary(m1_common)
```
# Observations from m1
For the full exploration sample (used in m1):
- n = 6075 apartments
- Distance coefficient = -0.594
- p < 2 x 10^-16
- R^2 = 0.0125
- In summary: an apartment located 100m farther from the nearest subway station is associated with about $59 lower monthly rent on average
For the complete-case exploration sample (used in m1_complete):
n = 1353 apartments
- Distance coefficient = -0.843
- p = 9.63 * 10^-10
- R^2 = 0.0273
- In summary: an apartment located 100m farther from the nearest subway station is associated with about $84 lower monthly rent on average

# Run m2

Model m2 adds key apartment characteristics such as number of bedrooms, number of bathrooms, square footage and borough to see if controlling for these variables provides a better explanation for our Y variable. 

```{r}


m2 <- lm(
  monthly_rent ~
    distance_to_subway_m +
    bedrooms +
    bathrooms +
    sqft +
    borough,
  data = model_df
)

summary(m2)
```
# Observations from m2

- Distance coefficient is -0.3675 so we can interpret this as about $36.75 lower rent per additional 100m of distance from nearest subway station
- Distance coefficient has p = 0.000662, providing strong statistical evidence of negative conditional association
- Adding control variables changed coefficient from -0.843 in m1 to -0.368 in m2
- The drastic change in the distance coefficient suggests that a substantial portion of the unconditional association between subway proximity and rent is explained by differences in observed apartment characteristics and borough. 
- Model dramatically improved - R^2 went from 0.0273 to 0.524

# Using neighborhood instead of borough

A potential model m3 could be just like m2 except we swap out borough for neighborhood as our main geographic control. There is massive variation within a borough (Williamsburg vs East New York in Brooklyn) so neighborhood might be a better indicator variable. 
```{r}
length(unique(model_df$neighborhood))
```
Results above show that there are 139 neighborhood groups, which is a lot of parameters relative to 1353 apartment observations. It is very likely that many of these neighborhoods will have only one or two listings, which will be useless in estimate coefficients. 

# Visualizing relationships

```{r}
ggplot(model_df, aes(x = distance_to_subway_m, y = monthly_rent)) +
  geom_point(alpha = 0.25) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    title = "Monthly Rent and Distance to Nearest Subway",
    x = "Distance to Nearest Subway (meters)",
    y = "Monthly Rent ($)"
  ) +
  theme_minimal()
```
```{r}
ggplot(model_df, aes(x = log_distance_m, y = log_monthly_rent)) +
  geom_point(alpha = 0.25) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    title = "Log Rent and Log Distance to Nearest Subway",
    x = "Log Distance to Nearest Subway",
    y = "Log Monthly Rent"
  ) +
  theme_minimal()
```
# Plot observations

Curve in the first plot drops steeply at relatively short distances from subway stops and then flattens considerably, meaning raw linear linear mapping is not the best form for our model. The log-log plot looks better behaved, with the scatter points being more spread out across the x-axis and the relationship looks more linear (though there is still a noticeable curve)

# Running m3

We'll actually try three different versions of m3 to see whether we need to transform one or both of the primary variables.

```{r}
# Linear rent, log distance
m3_logx <- lm(
  monthly_rent ~
    log_distance_m +
    bedrooms +
    bathrooms +
    sqft +
    borough,
  data = model_df
)

# Log rent, linear distance
m3_logy <- lm(
  log_monthly_rent ~
    distance_to_subway_m +
    bedrooms +
    bathrooms +
    sqft +
    borough,
  data = model_df
)

# Log rent, log distance
m3_loglog <- lm(
  log_monthly_rent ~
    log_distance_m +
    bedrooms +
    bathrooms +
    sqft +
    borough,
  data = model_df
)

# Exploring relationship between borough and 
m3_loglog_distance_borough <- lm(
  log_monthly_rent ~
    bedrooms +
    bathrooms +
    sqft +
    log_distance_m*borough,
  data = model_df
)

summary(m3_logx)
summary(m3_logy)
summary(m3_loglog)
summary(m3_loglog_distance_borough)
```
# Observations from m3

- When logging our X variable (distance), our R^2 value only goes up to 0.5233 from the earlier model's 0.5211 , so logging distance on its own doesn't get us very much - in this case, coefficient of -314.39 means that for 1% increase in distance associates with a $3.14 decrease in monthly rent
- When logging rent, the model's adjusted R^2 jumps to 0.5656, which indicates a better form for the model -in this case, coefficient of -0.00008089 means that for every 100m of additional difference, there is an associated 0.81% decrease in monthly rent when holding other characteristics constant
- Logging both variables gives us a R^2 of 0.5677, our highest yet (but not that much higher than logging just rent) - this version with its coefficient of -0.06296 means that 1% increase in distance is associated with approximately 0.063% decrease in monthly rent
- For fun, when we measure the interaction between borough and log_distance_m, we find that our R^2 value hardly changes (from 0.5677 to 0.5702) and our p-values for the interaction terms remain high - this means the relationship between subway distance and rent does not vary much when we change borough. 

# Creating residual vs fitted plots
```{r}

plot(m2, which = 1, main = "Level-Level")
plot(m3_logx, which = 1, main = "Log X")
plot(m3_logy, which = 1, main = "Log Y")
plot(m3_loglog, which = 1, main = "Log-Log")
```
# Plot observations

When we are logging neither variable or just logging distance, residual values fall below 0 and then rise considerably at higher fitted rents. Additionally, the vertical spread of residuals gets much larger as fitted rent increases. When we are logging just rent or logging both rent and distance, the residual cloud becomes more evenly distributed around zero, with a flatter line. Based off this, we can say it makes sense to use log_monthly_rent since raw rent is extremely right skewed and the original rent vs distance relationship was strongly nonlinear. 

# Evaluating heteroskedasticity
```{r}
coeftest(m3_logy, vcov = vcovHC(m3_logy, type = "HC1"))

coeftest(m3_loglog, vcov = vcovHC(m3_loglog, type = "HC1"))
```
Checking the robust standard errors does not change our substantive conclusion. In both the log Y and log log models, distance to the nearest subway station remains statistically and significantly associated with monthly rent (p < 0.001). In the log-log model, a 10% increase in distance from the subway is associated with a 0.63% decrease in monthly rent, holding apartment characteristics and borough constant. 

# Experimenting M4_neighborhood

We'll try swapping out borough for neighborhood to see if we get a better fitted model, keeping in mind the tradeoff we discussed earlier with the sheer number of neighborhood indicator variables. 

```{r}
m4_neighborhood <- lm(
  log_monthly_rent ~
    log_distance_m +
    bedrooms +
    bathrooms +
    sqft +
    neighborhood,
  data = model_df
)

coeftest(
  m4_neighborhood,
  vcov = vcovHC(m4_neighborhood, type = "HC1")
)

coef(m4_neighborhood)["log_distance_m"]

summary(m4_neighborhood)$adj.r.squared
```

# M4_neighborhood observations

- Our distance coefficient has shrunk from -0.063 to -0.015 - our estimated association has become 0.15% lower rent for 10% increase in subway distance when holding apartment characteristics and neighborhood constant
- Our distance coefficient is no longer statistically significant using robust standard errors (SE = 0.012, p = 0.219) - this means geographic confounding is huge
- Our R^2 has jumped from 0.568 to 0.760
- This could suggest much of the association observed in less geographically controlled models may reflect differences in rents and subway accessibility across neighborhoods rather than differences in subway proximity within neighborhoods.
- Trade-off here is that we are using 138 neighborhood indicators based on sparse groups. 

# Running M4_geographic_coordinates

We'll try to run a model that is using our latitude and longitude as a geography control

```{r}
m4_geo_coordinates <- lm(
  log_monthly_rent ~
    log_distance_m +
    bedrooms +
    bathrooms +
    sqft +
    latitude +
    longitude +
    borough,
  data = model_df
)

coeftest(
  m4_geo_coordinates,
  vcov = vcovHC(m4_geo, type = "HC1")
)

coef(m4_geo_coordinates)["log_distance_m"]

summary(m4_geo_coordinates)$adj.r.squared
```
#M4_geographic_coordinates observations

- Our distance coefficient becomes -0.02564, meaning 10% increase in distance from nearest subway station is associated with 0.26% decrease in monthly rent
- p-value is 0.0228, which shows statistical significance that the neighborhood model did not
- The change in distance coefficient from -0.063 in the log-log borough model to -0.026 in the log-log borough model that also controls for geographic coordinates implies that within borough geographic variation accounts for substantial portion of the observed association. 
- Our R^2 is 0.61 - better than in our log-log borough model but worse than our log-log neighborhood model
- Longitude and latitude having a linear relationship with rent is pretty crude though 
- Overall, this model independently points us in the same direction as the neighborhood model - more fine-grained location control shows location can be a significant confounder

# Pulling in our confirmation dataset

```{r}

# Full exploration sample
confirm_df <- modeling_data |>
  filter(analysis_split == "confirmation")

# Complete-case sample for models requiring all core covariates
confirm_model_df <- confirm_df |>
  drop_na(
    monthly_rent,
    bedrooms,
    bathrooms,
    sqft,
    borough,
    distance_to_subway_m
  )
```
```{r}

```
# Running m3log_log model on our confirmation dataset

```{r}
# Log rent, log distance
confirm_m3_loglog <- lm(
  log_monthly_rent ~
    log_distance_m +
    bedrooms +
    bathrooms +
    sqft +
    borough,
  data = confirm_model_df
)

summary(confirm_m3_loglog)
```
# Observations from m3_log_log on confirmation dataframe

- Logging both variables gives us a R^2 of 0.6164 and a distance coefficient of -0.07137 means that 1% increase in distance is associated with approximately 0.07137% decrease in monthly rent

# Creating residuals vs fitted plot for m3_log_log on confirmation dataframe

```{r}
plot(confirm_m3_loglog, which = 1, main = "Log-Log")
```
# Observations from residuals vs fitted plot of m3_log_log on confirmation dataframe

Our residual cloud is very evenly distributed around 0 in this model with a slight increase in vertical spread of the residuals as the logged fitted rent values becomes larger.

# Evaluating heteroskedasticity

```{r}
coeftest(confirm_m3_loglog, vcov = vcovHC(confirm_m3_loglog, type = "HC1"))
```
# Observations from t-test

Running the t-test shows that are our standard errors are fairly robust. In this log-log model, a 10% increase in distance from the subway is associated with a 0.71% decrease in monthly rent, holding apartment characteristics and borough constant.

```
